### 2024-07-24 chap_RL/chap_RL.tex
Grammarly done for the core of the text

To navigate through the myopic transitions not constrained by a prescribed \ce{CO2}-trajectory and investigate the efficiency of different policies, we apply the reinforcement learning approach detailed in Section \ref{sec:meth:RL} on the case study of Belgium. Besides the environment, \ie the myopic transition pathway of the whole-energy system via EnergyScope Pathway (see Chapter \ref{chap:chap_methodo}), the first part of this chapter presents the three key features of interaction between the agent, optimizing its policy, and the environment: actions, states and reward.  Then, the results of this policy optimisation point out strategies to follow, \ie \textit{sweet spots}, in the transitions under uncertainties as well as \textit{no-go zones} where the chances of succeeding the transition, \ie respecting the \ce{CO2} budget, are very limited. Finally, these results are compared with references, \ie the perfect foresight and the myopic optimisation of the transition under the same uncertainties but without the trained \gls{RL}-agent that can support this transition thanks to its learned policy.

@@

Comparing the \gls{RL}-based results with more conventional approaches, \ie perfect foresight optimisation, highlights the added-value brought by this approach in the exploration of myopic transition pathways.

@@

\newpage
\section{Comparison with references}
\label{sec:RL:testing}
After investigating the exploration of myopic transitions, this section compares these results with the perfect foresight under uncertainties from the \gls{GSA} presented in Chapter \ref{chap:atom_mol}. To reduce the computational time, the monthly model could also be used to carry out \gls{RL}-based exploration of myopic pathways. Appendix \ref{app:app_RL_TD_MO} presents the comparison between the results obtained from the learning process on the hourly and monthly models.

\gls{RL}-based myopic optimisation provides \ce{CO2}-emissions pathways different from the perfect foresight approach to respect the same \ce{CO2} budget (see left side of Figure \ref{fig:Gwp_pathway_total_tran_cost}). However, driven first by this \ce{CO2} budget, the agent often reaches much lower cumulative emissions when succeeding the transition (see Figure \ref{fig:Cum_gwp_cost}). This comes from the agent's actions that limit the emissions and/or the consumption of fossil resources at the early stages. Thanks to the bigger reduction of emissions at these early stages, the \gls{RL}-based optimisation can benefit from a ``\ce{CO2}-buffer'' at the end of the transition. This buffer is compensated by the end of the transition where 50\% of the myopic transitions reach 2050 with 10 or more remaining Mt$_{\ce{CO2},\text{eq}}$ compared to 4 for the perfect foresight approach. These remaining emissions by 2050 come from the consumption in industrial boilers of waste and coal that account for 3.5\% and 2.4\% of the primary mix on average by 2050. Finally, the long-term vision of the perfect foresight approach results in smoother reduction of the emissions to end up with less emissions by 2050.

The comparison between the failures and the successes demonstrates the added-value brought by myopic pathway optimisation. In the near term (2025-2030), levels of emission are similar between perfect foresight and myopic cases that have failed. This shows that limited foresight encourages to strongly act at the early stages. On top of this, following the initial steps of \ce{CO2}-emissions pathways resulting from PF approach would likely ($\sim$80\%) lead to fail the transition.

Looking at the total transition cost, the combination of the agent's actions and favourable economic conditions (see Section \ref{subsec:RL:learning:states}) make the myopic transitions cheaper, on average, than the PF cases (see right side of Figure \ref{fig:Gwp_pathway_total_tran_cost}). This is also due to the fact that the perfect foresight approach always finds a solution even in worst conditions such as high cost of purchasing resources and high \gls{EUD}. This explains the wider variability of the PF results too. However, with the same sample of uncertain parameters, given the assumed full knowledge over the whole time horizon, PF naturally results in a cheaper transition than its myopic equivalent.

\begin{figure}[!htbp]
\centering
\includegraphics[height=3.5cm]{Gwp_pathway_core.pdf}
\includegraphics[height=3.5cm]{Transition_cost_comp_2.pdf}
\caption{Comparison of \ce{CO2}-emissions pathways (left) and total transition cost (right) from the perfect foresight optimisation under uncertainties and the \gls{RL}-based myopic optimisation. Myopic transitions succeed with a more drastic reduction of emissions in the short-term and, on average, more favourable economic conditions.}
\label{fig:Gwp_pathway_total_tran_cost}
\end{figure}

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.325\textwidth]{Opex_2050_comp_core.pdf}
\includegraphics[width=0.325\textwidth]{Capex_2050_comp_core.pdf}
\includegraphics[width=0.325\textwidth]{Salvage_2050_comp_core.pdf}
\caption{Comparison of cumulative OPEX (left), CAPEX (center) and salvage value (right) in 2050 from the perfect foresight optimisation under uncertainties and the \gls{RL}-based myopic optimisation.}
\label{fig:Opex_Capex_Salvage_comp}
\end{figure}

The cost of purchasing the energy carriers represents about 70\% of the total cumulative \gls{OPEX}. The assessment of the primary energy mix by 2050 highlights that the difference of OPEX between the perfect foresight and the myopic pathways come from the import of electrofuels, and especially of e-ammonia (see Figure \ref{fig:Mix_2050_comp}).  In the majority of the cases, e-ammonia is more than two times more imported in the myopic transitions. Being cheaper than e-methane (see Chapter \ref{chap:case_study}), e-ammonia brings flexibility in the production of electricity via \gls{CCGT} (see Chapter \ref{chap:atom_mol}). Besides the slightly favourable economical conditions (see Table \ref{tab:param_RL}), the myopic optimisations opt to invest massively into importing renewable molecules because of the limited knowledge in the future, and, among others, the availability of \gls{SMR}. This explains why 50\% of the successful transitions reached cumulative emissions below 900\,Mt$_{\ce{CO2},\text{eq}}$ (see Figure \ref{fig:Cum_gwp_cost}).

@@

\section{Conclusions}
\label{sec:RL:conclusions}
In the literature, two options are investigated to explore transition pathways of a whole-energy system: perfect foresight and myopic. In perfect foresight optimisation, a full knowledge of the parameters is assumed over the entire time horizon. To add more realism, myopic optimisation considers a sequence of more limited time windows leading to the end of the time horizon \cite{poncelet2016myopic}. To respect a \ce{CO2} budget over the transition, this case requires a prescribed \ce{CO2}-trajectory \cite{fais2016impact}. However, since the effect of \ce{CO2}-emissions in climate change is cumulative, the total amount of these emissions matters more than the trajectory itself. Consequently, we have applied the \acrfull{RL} approach on the environment of the Belgian myopic pathway optimisation under uncertainties.

@@

This \gls{RL}-based exploration pointed out that short-term actions were needed to hope succeeding such a transition, as also demonstrated by \citet{luderer2018residual}. Where \gls{LFO} becomes less cost-competitive in the near-future, limiting the use of coal should be done at any cost. Then, fossil gas should be replaced by e-methane in the mid-term while putting a strict limit on the overall emissions becomes the most effective action by the end of the transition. The analysis of the share of renewables in the primary energy mix highlighted intermediate milestones to have higher chances to succeed the transition. Below 54\% of renewables in the mix in the near-future, these chances become much more limited, \ie the no-go zone.

We have compared the results coming from this \gls{RL}-based myopic optimisation with the hourly perfect foresight approach. These myopic optimisations provided pathways respecting the \ce{CO2} budget that were more drastic in cutting emissions in the short-term than the perfect. Supported by the agent's actions, we could reach cheaper energy system than what the perfect foresight could give. This comparative analysis also pointed out that investing more into renewable electrofuels was the option when knowledge about the future is limited. To do so, it requires reducing the uncertainties of their cost of purchasing.

@@

In conclusion, via the application of the \gls{RL} approach, we have widely explored the different myopic transition pathways and identified sweet-spots (and no-go zones) to succeed a transition with an ambitious \ce{CO2} budget target. It also highlighted the actions to take to effectively support a whole-energy system transition. This framework allows future research to test other policies on other case studies and identify the key actions and the timing to take them through the transition.

>>>>

To navigate through the myopic transitions not constrained by a prescribed \ce{CO2} trajectory and investigate the efficiency of different policies, we apply the reinforcement learning approach detailed in Section \ref{sec:meth:RL} on the case study of Belgium. Besides the environment, \ie the myopic transition pathway of the whole-energy system via EnergyScope Pathway (see Chapter \ref{chap:chap_methodo}), the first part of this chapter presents the three key features of interaction between the agent, optimizing its policy, and the environment: actions, states and reward.  Then, the results of this policy optimisation point out strategies to follow, \ie \textit{sweet spots}, in the transitions under uncertainties as well as \textit{no-go zones} where the chances of succeeding the transition, \ie respecting the \ce{CO2} budget, are very limited. Finally, these results are compared with references, \ie the perfect foresight and the myopic optimisation of the transition under the same uncertainties but without the trained \gls{RL}-agent that can support this transition thanks to its learned policy.

@@

Comparing the \gls{RL}-based results with more conventional approaches, \ie perfect foresight optimisation, highlights the added value brought by this approach in the exploration of myopic transition pathways.

@@

\newpage
\section{Comparison with references}
\label{sec:RL:testing}
After investigating the exploration of myopic transitions, this section compares these results with the perfect foresight under uncertainties from the \gls{GSA} presented in Chapter \ref{chap:atom_mol}. To reduce the computational time, the monthly model could also be used to carry out \gls{RL}-based exploration of myopic pathways. Appendix \ref{app:app_RL_TD_MO} compares the results obtained from the learning process on the hourly and monthly models.

\gls{RL}-based myopic optimisation provides \ce{CO2} emissions pathways different from the perfect foresight approach to respect the same \ce{CO2} budget (see left side of Figure \ref{fig:Gwp_pathway_total_tran_cost}). However, driven first by this \ce{CO2} budget, the agent often reaches much lower cumulative emissions when succeeding the transition (see Figure \ref{fig:Cum_gwp_cost}). This comes from the agent's actions that limit the emissions and/or the consumption of fossil resources at the early stages. Thanks to the bigger emission reduction at these early stages, the \gls{RL}-based optimisation can benefit from a ``\ce{CO2} buffer'' at the end of the transition. This buffer is compensated by the end of the transition where 50\% of the myopic transitions reach 2050 with 10 or more remaining Mt$_{\ce{CO2},\text{eq}}$ compared to 4 for the perfect foresight approach. These remaining emissions by 2050 come from the consumption in industrial boilers of waste and coal accounting for 3.5\% and 2.4\% of the primary mix on average by 2050. Finally, the long-term vision of the perfect foresight approach results in a smoother reduction of emissions to end up with less emissions by 2050.

The comparison between the failures and the successes demonstrates the added value brought by myopic pathway optimisation. In the near term (2025-2030), levels of emission are similar between perfect foresight and myopic cases that have failed. This shows that limited foresight encourages to strongly act at the early stages. On top of this, following the initial steps of \ce{CO2} emissions pathways resulting from the PF approach would likely ($\sim$80\%) lead to failure of the transition.

Looking at the total transition cost, the combination of the agent's actions and favourable economic conditions (see Section \ref{subsec:RL:learning:states}) make the myopic transitions cheaper, on average, than the PF cases (see right side of Figure \ref{fig:Gwp_pathway_total_tran_cost}). This is also because the perfect foresight approach always finds a solution even in the worst conditions such as the high cost of purchasing resources and high \gls{EUD}. This explains the wider variability of the PF results too. However, with the same sample of uncertain parameters, given the assumed full knowledge over the whole time horizon, PF naturally results in a cheaper transition than its myopic equivalent.

\begin{figure}[!htbp]
\centering
\includegraphics[height=3.5cm]{Gwp_pathway_core.pdf}
\includegraphics[height=3.5cm]{Transition_cost_comp_2.pdf}
\caption{Comparison of \ce{CO2} emissions pathways (left) and total transition cost (right) from the perfect foresight optimisation under uncertainties and the \gls{RL}-based myopic optimisation. Myopic transitions succeed with a more drastic reduction of emissions in the short term and, on average, more favourable economic conditions.}
\label{fig:Gwp_pathway_total_tran_cost}
\end{figure}

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.325\textwidth]{Opex_2050_comp_core.pdf}
\includegraphics[width=0.325\textwidth]{Capex_2050_comp_core.pdf}
\includegraphics[width=0.325\textwidth]{Salvage_2050_comp_core.pdf}
\caption{Comparison of cumulative OPEX (left), CAPEX (centre) and salvage value (right) in 2050 from the perfect foresight optimisation under uncertainties and the \gls{RL}-based myopic optimisation.}
\label{fig:Opex_Capex_Salvage_comp}
\end{figure}

The cost of purchasing the energy carriers represents about 70\% of the total cumulative \gls{OPEX}. The assessment of the primary energy mix by 2050 highlights that the difference in OPEX between the perfect foresight and the myopic pathways comes from the import of electrofuels, and especially of e-ammonia (see Figure \ref{fig:Mix_2050_comp}).  In the majority of the cases, e-ammonia is more than two times more imported in the myopic transitions. Being cheaper than e-methane (see Chapter \ref{chap:case_study}), e-ammonia brings flexibility in the production of electricity via \gls{CCGT} (see Chapter \ref{chap:atom_mol}). Besides the slightly favourable economic conditions (see Table \ref{tab:param_RL}), the myopic optimisations opt to invest massively into importing renewable molecules because of the limited knowledge of the future, and, among others, the availability of \gls{SMR}. This explains why 50\% of the successful transitions reached cumulative emissions below 900\,Mt$_{\ce{CO2},\text{eq}}$ (see Figure \ref{fig:Cum_gwp_cost}).

@@

\section{Conclusions}
\label{sec:RL:conclusions}
In the literature, two options are investigated to explore transition pathways of a whole-energy system: perfect foresight and myopic. In perfect foresight optimisation, a full knowledge of the parameters is assumed over the entire time horizon. To add more realism, myopic optimisation considers a sequence of more limited time windows leading to the end of the time horizon \cite{poncelet2016myopic}. To respect a \ce{CO2} budget over the transition, this case requires a prescribed \ce{CO2} trajectory \cite{fais2016impact}. However, since the effect of \ce{CO2} emissions in climate change is cumulative, the total amount of these emissions matters more than the trajectory itself. Consequently, we have applied the \acrfull{RL} approach on the environment of the Belgian myopic pathway optimisation under uncertainties.

@@

This \gls{RL}-based exploration pointed out that short-term actions were needed to hope to succeed such a transition, as also demonstrated by \citet{luderer2018residual}. Where \gls{LFO} becomes less cost-competitive in the near future, limiting the use of coal should be done at any cost. Then, fossil gas should be replaced by e-methane in the mid-term while putting a strict limit on the overall emissions becomes the most effective action by the end of the transition. The analysis of the share of renewables in the primary energy mix highlighted intermediate milestones to have higher chances of succeeding the transition. Below 54\% of renewables in the mix in the near future, these chances become much more limited, \ie the no-go zone.

We have compared the results coming from this \gls{RL}-based myopic optimisation with the hourly perfect foresight approach. These myopic optimisations provided pathways respecting the \ce{CO2} budget that were more drastic in cutting emissions in the short term than the perfect foresight. This comparative analysis also pointed out that investing more in renewable electrofuels was the option when knowledge about the future is limited. To do so, it requires reducing the uncertainties of their cost of purchasing.

@@

In conclusion, via the application of the \gls{RL} approach, we have widely explored the different myopic transition pathways and identified sweet spots (and no-go zones) to succeed a transition with an ambitious \ce{CO2} budget target. It also highlighted the actions to take to effectively support a whole-energy system transition. This framework allows future research to test other policies on other case studies and identify the key actions and the timing to take them through the transition.
